Local PCA Shows How the Effect of Population Structure Differs Along the Genome

### 2016-09-25 writeup/local_pca_paper.tex
stuff for submission

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}

@@

These realized patterns of genetic relatedness
summarize the shapes of the genealogical trees % that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does relatedness,
but averaging over sufficiently many trees we hope to get a stable estimate
that doesn't depend much on the genetic markers chosen.
This is not guaranteed:
for instance,
relatedness on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{kim2002hitchhiking,charlesworth1993effect,barton2000genetic}.
Indeed,
many species show chromosome-scale variation in diversity and divergence
\citep[e.g.,][]{langley2012genomic};
species phylogenies can differ along the genome 
due to incomplete lineage sorting,
adaptive introgression and/or local adaptation 
\citep[e.g.,][]{pease2013accurate,ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting};
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citep{charlesworth2003review}.

Patterns in genome-wide relatedness are often summarized
by applying principal components analysis \citep[PCA,][]{patterson2006population} 
to the genetic covariance matrix,
as pioneered by \citet{menozzi1978synthetic}.
% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.
The results of PCA can be related to the genealogical history of the samples, 
such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}, 
and sometimes produce ``maps'' of population structure
that reflect the samples' geographic origin distorted by rates of gene flow
\citep{novembre2008genes}.
% although other patterns can be caused by recent migration or nongeographic kinship patterns.  % \citep{astle2009population}.
% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

Modeling such ``background'' kinship between samples
is essential to genome-wide association studies \citep[GWAS,][]{price2006principal,astle2009population},
% and PCA has often been used in this context \citep{price2006principal}.
and so understanding variation in kinship along the genome 
% can help us to have a better understanding of the relation between genome structure and population structure, 
could lead to more generally powerful methods,
and may be essential for doing GWAS in species with substantial heterogeneity in realized patterns of mean relatedness along the genome.

@@

A note on nomenclature: In this work we describe variation in patterns of relatedness using local PCA,
where ``local'' refers to proximity along the genome.
A number of general methods for dimensionality reduction also use a strategy of ``local PCA''
\citep[e.g.,][]{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear},
performing PCA not on the entire dataset but instead on subsets of observations,
providing local pictures which are then stitched back together to give a global picture.
At first sight,
this differs from our method in that we restrict to subsets of \emph{variables} instead of subsets of observations.
However, if we flip perspectives and think of each genetic variant as an observation,
our method shares common threads, although 
our method does not subsequently use adjacency along the genome,
as we aim to identify similar regions that may be distant.
% Future methods for visualization of genomic data
% may benefit from other advances in this substantial literature 
% \citep[reviewed in][]{vandermaaten2009dimensionality}.

@@

\begin{figure}
    \begin{center}
       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
    \end{center}
    \caption{
         As in Figure 2, except that samples are colored by orientation of the polymorphic inversions 
         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
         \citep[data from][]{lack2015drosophila}
         In each ``INV'' denotes an inverted genotype, ``ST'' denotes the standard orientation,
         and ``N'' denotes unknown.
        \label{fig:pca_by_inversion}
    }
\end{figure}

@@

\paragraph{\textit{Drosophila melanogaster}:}
We used whole-genome sequencing data 
from the Drosophila Genome Nexus \citep[\url{http://www.johnpool.net/genomes.html},][]{lack2015drosophila},
consisting of the Drosophila Population Genomics Project phases 1--3 \citep{langley2012genomic,pool2012population},
and additional African genomes \citep{lack2015drosophila}.
After removing 20 genomes with more than 8\% missing data,
we were left with 380 samples from 16 countries across Africa and Europe.
Since the \textit{Drosophila} samples are from inbred lines or haploid embryos, 
we treat the samples as haploid when recoding;
regions with residual heterozygosity were marked as missing in the original dataset;
we also removed positions with more than 20\% missing data. 
% (We chose these cutoffs as the tails of the relevant empirical distributions.)
Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs;
PCA plots for each arm are shown in Figure \ref{sfig:pca_drosophila_allchr}.

@@

\bibliographystyle{plainnat}
\bibliography{references}

>>>>

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{setspace}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}

% % stuff for submission
% \usepackage[nomarkers,figuresonly]{endfloat}
% \usepackage{setspace}
% \doublespacing
% \usepackage{lineno}
% \linenumbers
% \bibliographystyle{plos2015}
% \newcommand\citet{\cite}
% \newcommand\citep{\cite}

% % for the nice version
\usepackage[square,sort,comma,numbers]{natbib} \bibpunct{(}{)}{;}{author-year}{}{,} 
\bibliographystyle{plainnat}

@@

These realized patterns of genetic relatedness
summarize the shapes of the genealogical trees % that relate the samples
at each location along the genome.
Since these trees vary along the genome, so does relatedness,
but averaging over sufficiently many trees we hope to get a stable estimate
that doesn't depend much on the genetic markers chosen.
This is not guaranteed:
for instance,
relatedness on sex chromosomes is expected to differ from the autosomes;
and positive or negative selection on particular loci can dramatically disort shapes of nearby genealogies
\citep{kim2002hitchhiking,charlesworth1993effect,barton2000genetic}.
Indeed,
many species show chromosome-scale variation in diversity and divergence
(e.g., \citep{langley2012genomic});
species phylogenies can differ along the genome 
due to incomplete lineage sorting,
adaptive introgression and/or local adaptation 
(e.g., \citep{pease2013accurate,ellegren2012genomic,nadeau2012genomic,pool2015natural,vernot2014resurrecting});
and theoretical expectations predict that geographic patterns of relatedness should depend on selection
\citep{charlesworth2003review}.

Patterns in genome-wide relatedness are often summarized
by applying principal components analysis (PCA, \citep{patterson2006population}) 
to the genetic covariance matrix,
as pioneered by \citet{menozzi1978synthetic}.
% About 37 years ago, \citet{menozzi1978synthetic} first applied principal component analysis (PCA) in population genetics to construct maps summarizing genetic variation \citep{menozzi1978synthetic}. 
% Nowadays, PCA is a widely used powerful non-parametric method to extract information from genetic data.
The results of PCA can be related to the genealogical history of the samples, 
such as time to most recent common ancestor and migration rate between populations \citep{novembre2008interpreting,mcvean2009genealogical}, 
and sometimes produce ``maps'' of population structure
that reflect the samples' geographic origin distorted by rates of gene flow
\citep{novembre2008genes}.
% although other patterns can be caused by recent migration or nongeographic kinship patterns.  % \citep{astle2009population}.
% Through dimension-reduction, PCA can identify key components of population structure, which describes how different samples are related, and are often closely related to geography. 
% Plots of the first two principal components (PCs) can mimic the samples' geographic origin to some extent. 
% Since population structure describes how different samples are related, samples living closer tend to be more genetically similar and thus tend to be clustered in PC plots \citep{novembre2008genes,patterson2006population}. 
% However this relatedness is limited while there's recent migration or for group with nongeographic kinship patterns,for example, social or religious groups. \citep{astle2009population}

Modeling such ``background'' kinship between samples
is essential to genome-wide association studies (GWAS, \citep{price2006principal,astle2009population}),
% and PCA has often been used in this context \citep{price2006principal}.
and so understanding variation in kinship along the genome 
% can help us to have a better understanding of the relation between genome structure and population structure, 
could lead to more generally powerful methods,
and may be essential for doing GWAS in species with substantial heterogeneity in realized patterns of mean relatedness along the genome.

@@

A note on nomenclature: In this work we describe variation in patterns of relatedness using local PCA,
where ``local'' refers to proximity along the genome.
A number of general methods for dimensionality reduction also use a strategy of ``local PCA''
(e.g., \citep{manjon2013diffusion,kambhatla1997dimension,weingessel2000local,roweis2000nonlinear}),
performing PCA not on the entire dataset but instead on subsets of observations,
providing local pictures which are then stitched back together to give a global picture.
At first sight,
this differs from our method in that we restrict to subsets of \emph{variables} instead of subsets of observations.
However, if we flip perspectives and think of each genetic variant as an observation,
our method shares common threads, although 
our method does not subsequently use adjacency along the genome,
as we aim to identify similar regions that may be distant.
% Future methods for visualization of genomic data
% may benefit from other advances in this substantial literature 
% (reviewed in \citep{vandermaaten2009dimensionality}).

@@

\begin{figure}
    \begin{center}
       \includegraphics{Fig3_pca_plots_color_by_inv_allchr_with_ChrX_In1A}
    \end{center}
    \caption{
         As in Figure 2, except that samples are colored by orientation of the polymorphic inversions 
         In(2L)t, In(2R)NS, In(3L)OK, In(3R)K and In(1)A respectively. 
         (data from \citep{lack2015drosophila})
         In each ``INV'' denotes an inverted genotype, ``ST'' denotes the standard orientation,
         and ``N'' denotes unknown.
        \label{fig:pca_by_inversion}
    }
\end{figure}

@@

\paragraph{\textit{Drosophila melanogaster}:}
We used whole-genome sequencing data 
from the Drosophila Genome Nexus (\url{http://www.johnpool.net/genomes.html}, \citep{lack2015drosophila}),
consisting of the Drosophila Population Genomics Project phases 1--3 \citep{langley2012genomic,pool2012population},
and additional African genomes \citep{lack2015drosophila}.
After removing 20 genomes with more than 8\% missing data,
we were left with 380 samples from 16 countries across Africa and Europe.
Since the \textit{Drosophila} samples are from inbred lines or haploid embryos, 
we treat the samples as haploid when recoding;
regions with residual heterozygosity were marked as missing in the original dataset;
we also removed positions with more than 20\% missing data. 
% (We chose these cutoffs as the tails of the relevant empirical distributions.)
Each chromosome arm we investigated (X, 2L, 2R, 3L, and 3R) has 2--3 million SNPs;
PCA plots for each arm are shown in Figure \ref{sfig:pca_drosophila_allchr}.

@@

\bibliography{references}
